Global search algorithm with local tuning for solving mixed-integer global optimization problems \thanks{This study was supported by the .}

### 2021-04-05 2021/MOTOR_BarkalovLebedev/MOTOR2021.tex
Правки MOTOR

Having several objective functions in MOO problems that determine the effectiveness of the decisions to be made results in most cases in situations where no option can be found in the whole set of alternatives that would be the best one with respect to all efficiency criteria at the same time. In such situations, the solution of MOO problems usually consists in finding non-dominated (or efficient) decisions in which values cannot be improved simultaneously across all objective functions. The set of all efficient decisions is usually referred to as the Pareto set.

Finding the whole Pareto set when solving MOO problems can be computationally intensive (especially when objective functions are multiextremal). Moreover, having the whole Pareto set may be redundant for the decision maker (DM) due to the complexity of analyzing a large number of possible alternatives. As a result, when solving MOO problems in practice, it may be sufficient to find one or several efficient decisions taking into account the DM's perceptions of the required optimality of the options to be chosen.

@@

One common property of methods based on scalarization of the vector efficiency criterion is the existence of some scalarization coefficients that can be varied to obtain different solutions from the Pareto set. Thus, the scalarization coefficients may be interpreted as measures of the importance of the efficiency criteria determined by DM according to his/her perception of the required optimality of the decisions to be made. As a result, the general scheme for solving the MOO problem can be represented as a sequence of steps; at each step, DM sets the necessary scalarization coefficients, then the resulting scalar optimization problem is solved, after that DM analyzes the efficient decision found and, if necessary, the above steps are repeated.

@@

In the most general form, the multi-objective optimization (MOO) problem can be formulated as follows
\begin{equation}
\label{eq:01}
f(y) = (f_1(y), f_2(y), \dots, f_s(y)) \to \min,  y\in D,
\end{equation}
where $f(y) = (f_1(y), f_2(y), \dots, f_s(y))$ are objective functions (efficiency criteria), $y = (y_1, y_2, \dots , y_N)$ is the vector of varied parameters, and $N$ is the dimensionality of the multi-objective optimization problem to be solved. The set of possible parameter values (search domain) $D$ is usually an $N$-dimensional hypercube
\begin{equation}
\label{eq:02}
D  = \{ y\in R^N: a_i \leq y_i \leq b_i, 1 \leq i \leq N \}
\end{equation}
for given boundary vectors $a$ and $b$.

Without loss of generality, it is assumed that objective functions should be minimized to improve the decision efficiency $y \in D$. It is also assumed that $f_i(y)$, $1 \leq i \leq s$, are multiextremal and have the form of time-consuming ``black-box'' computational procedures. It is also assumed that the objective functions $f_i(y)$, $1 \leq i \leq s$, satisfy the Lipschitz condition 
\begin{equation}
\label{eq:03}
|f_i(y') - f_i(y'')| \leq L_i\|y' - y''\|, y', y'' \in D, 1 \leq i \leq s,
\end{equation}
where $L_i$, $1 \leq i \leq s$, are the Lipschitz constants and $\|*\|$ denotes the Euclidean norm in $R^N$.  Condition (\ref{eq:03}) means that for small variations of the parameter $y \in D$, the corresponding changes in the values of the functions $f_i(y)$, $1 \leq i \leq s$, are bounded.

@@

\textbf{1. Scalarization of multiple objective functions.} In the most general form, the global optimization problem generated by the scalarization of multiple objective functions of the MOO problem can be represented in the form
\begin{equation}
\label{eq:04}
\min_{y \in D} {\varphi(y)} = \min_{y \in D} F(\lambda, y),
\end{equation}
where $F$ is a scalar multiextremal function generated as a result of scalarization of objective functions $f_i$, $1 \leq i \leq s$, $\lambda$ is the vector of parameters of the applied convolution of functions, and $D$ is the search domain from (\ref{eq:01}). In the proposed approach, we use for scalarization the compromise programming method \cite{c2,c11}, where solution of the MOO problem consists in finding the efficient decision corresponding most closely to the values of the objective functions of the specified reference decision $y^0 \in D$. In this case, a possible statement of scalar optimization problem can have the form 
\begin{equation}
\label{eq:05}
\min_{y \in D} F(\lambda,y)=\min_{y \in D} \left\{ \frac{1}{s} \sum_{i=1}^s {\lambda_i (f_i (y)-f_i (y^0 ))^2} \right\},
\end{equation}
where $F(\lambda,y)$ is the standard deviation of the values of the objective functions $f_i$, $1 \leq i \leq s$, for the decision $y \in D$ and for the specified reference decision $y^0 \in D$, while the coefficients $\lambda_i$, $0 \leq \lambda_i \leq 1$, $1 \leq i \leq s$, are measures of the significance of the approximation accuracy for each variable $y_i$, $1 \leq i \leq N$, separately. Without loss of generality, we can assume that the area of possible values of the coefficients $\lambda$ is a set 
\begin{equation}
\label{eq:06}
\lambda=(\lambda_1,\lambda_2, \dots, \lambda_s)\in \Lambda \subset R^s: \sum_{i=1}^{s}\lambda_i=1, \lambda_i \geq 0, 1 \leq i \leq s.
\end{equation}

The reference decision $y^0 \in D$ in (\ref{eq:05}) can be known \textit{a priori} or determined on the basis of some known prototype. In many cases, an abstract ideal decision $y^0 \in D$ is used as the reference decision $y^*\in D$ in which the objective functions $f_i$, $1 \leq i \leq s$, have minimal possible values, i.e.
\begin{equation}
\label{eq:07}
f_i^* =f_i (y^*)=\min_{y \in D} f_i (y), 1 \leq i \leq s.
\end{equation}

@@

\textbf{2. Dimensionality reduction.} As already mentioned, problem (\ref{eq:04}) is a multidimensional global optimization problem. Problems of this kind are computationally complex and are known to be subject to the ``curse of dimensionality'' -- computational complexity increases exponentially with increasing dimensionality of the optimization problem being solved \cite{c13,c14,c15,c16,c17,c18,c19,c20}. Nevertheless, the computational complexity of global optimization algorithms can be significantly reduced by dimensionality reduction based on the use of a Peano space-filling curve (or evolvent) $y(x)$ that uniquely and continuously maps the segment $[0,1]$ on an $N$-dimensional hypercube $D$ -- see, for example, \cite{c15,c21}. As a result of such reduction, multidimensional global optimization problems (\ref{eq:04}) are reduced to one-dimensional problems
\begin{equation}
\label{eq:09}
\min_{x \in [0,1]} \varphi (y(x))= \min_{y \in D} \varphi(y).
\end{equation}

@@

\textbf{3. Solving the one-dimensional reduced optimization problem.} Using the dimensionality reduction results in one more additional advantage of the proposed approach: many well-known one-dimensional global search algorithms (possibly, after some additional generalisation) can be used to solve the initial multidimensional MOO problem from (\ref{eq:01}) \cite{c22,c23,c24,c25,c26,c27,c28,c29}. At the same time, it should be noted that most works where dimensionality reduction is a key feature for solving multiextremal problems like (\ref{eq:04}) rely on the information-statistical theory of global search \cite{c15}. This theory has provided the a basis for developing a large number of efficient methods for multi-extremal optimization \cite{c28,c29,c30,c31,c32,c33,c34,c35,c36,c37}.

>>>>

@@

One common property of methods based on scalarization of the vector efficiency criterion is the existence of some scalarization coefficients that can be varied to obtain different solutions from the Pareto set. Thus, the scalarization coefficients may be interpreted as measures of the importance of the efficiency criteria determined by the decision maker (DM) according to his/her perception of the required optimality of the decisions to be made. As a result, the general scheme for solving the MOO problem can be represented as a sequence of steps; at each step, DM sets the necessary scalarization coefficients, then the resulting scalar optimization problem is solved, after that DM analyzes the efficient decision found and, if necessary, the above steps are repeated.

@@

In the most general form, the MCO problem can be formulated as follows
\begin{equation}
\label{eq:01}
f(y) = (f_1(y), f_2(y), \dots, f_s(y)) \to \min,  y\in D,
\end{equation}
where $f(y) = (f_1(y), f_2(y), \dots, f_s(y))$ are objective functions (efficiency criteria), $y = (y_1, y_2, \dots , y_N)$ is the vector of varied parameters, and $N$ is the dimensionality of the multi-objective optimization problem to be solved. The set of possible parameter values (search domain) $D$ is usually an $N$-dimensional hyperinterval
\begin{equation}
\label{eq:02}
D  = \{ y\in R^N: a_i \leq y_i \leq b_i, 1 \leq i \leq N \}
\end{equation}
for given boundary vectors $a$ and $b$.

Without loss of generality, it is assumed that objective functions should be minimized to improve the decision efficiency $y \in D$. It is also assumed that $f_i(y)$, $1 \leq i \leq s$, are multiextremal and have the form of time-consuming ``black-box'' computational procedures. It is also assumed that the objective functions $f_i(y)$, $1 \leq i \leq s$, satisfy the Lipschitz condition 
\begin{equation}
\label{eq:03}
|f_i(y') - f_i(y'')| \leq L_i\|y' - y''\|, y', y'' \in D, 1 \leq i \leq s,
\end{equation}
where $L_i$, $1 \leq i \leq s$, are the Lipschitz constants and $\|*\|$ denotes the Euclidean norm in $R^N$.  Condition (\ref{eq:03}) means that for variations of the parameter $y \in D$, the corresponding changes in the values of the functions $f_i(y)$, $1 \leq i \leq s$, are bounded.

@@

\subsection{Scalarization of multiple objective functions.} 
In the most general form, the global optimization problem generated by the scalarization of multiple objective functions of the MOO problem can be represented in the form
\begin{equation}
\label{eq:04}
\min_{y \in D} {\varphi(y)} = \min_{y \in D} F(\lambda, y),
\end{equation}
where $F$ is a scalar multiextremal function generated as a result of scalarization of objective functions $f_i$, $1 \leq i \leq s$, $\lambda$ is the vector of parameters of the applied convolution of functions, and $D$ is the search domain from (\ref{eq:01}). In the proposed approach, we use for scalarization the compromise programming method \cite{c2,c11}, where solution of the MOO problem consists in finding the efficient decision corresponding most closely to the values of the objective functions of the specified reference decision $y^0 \in D$. In this case, a possible statement of scalar optimization problem can have the form 
\begin{equation}
\label{eq:05}
\min_{y \in D} F(\lambda,y)=\min_{y \in D} \left\{ \frac{1}{s} \sum_{i=1}^s {\lambda_i (f_i (y)-f_i (y^0 ))^2} \right\},
\end{equation}
where $F(\lambda,y)$ is the standard deviation of the values of the objective functions $f_i$, $1 \leq i \leq s$, for the decision $y \in D$ and for the specified reference decision $y^0 \in D$, while the coefficients $\lambda_i$, $0 \leq \lambda_i \leq 1$, $1 \leq i \leq s$, are measures of the significance of the approximation accuracy for each variable $y_i$, $1 \leq i \leq N$, separately. Without loss of generality, we can assume that the area of possible values of the coefficients $\lambda$ is a set 
\begin{equation}
\label{eq:06}
\lambda=(\lambda_1,\lambda_2, \dots, \lambda_s)\in \Lambda \subset R^s: \sum_{i=1}^{s}\lambda_i=1, \lambda_i \geq 0, 1 \leq i \leq s.
\end{equation}

The reference decision $y^0 \in D$ in (\ref{eq:05}) can be known \textit{a priori} or determined on a basis of some known prototype. In many cases, an abstract ideal decision $y^0 \in D$ is used as the reference decision $y^*\in D$ in which the objective functions $f_i$, $1 \leq i \leq s$, have minimal possible values, i.e.
\begin{equation}
\label{eq:07}
f_i^* =f_i (y^*)=\min_{y \in D} f_i (y), 1 \leq i \leq s.
\end{equation}

@@

\subsection{Dimensionality reduction.} 
As already mentioned, problem (\ref{eq:04}) is a multidimensional global optimization problem. Problems of this kind are computationally complex and are known to be subject to the ``curse of dimensionality'' -- computational complexity increases exponentially with increasing dimensionality of the optimization problem being solved \cite{c13,c14,c15,c16,c17,c18,c19,c20}. Nevertheless, the computational complexity of global optimization algorithms can be significantly reduced by dimensionality reduction based on the use of a Peano space-filling curve (or evolvent) $y(x)$ that uniquely and continuously maps the segment $[0,1]$ on an $N$-dimensional hypercube $D$ -- see, for example, \cite{c15,c21}. As a result of such reduction, multidimensional global optimization problems (\ref{eq:04}) are reduced to one-dimensional problems
\begin{equation}
\label{eq:09}
\min_{x \in [0,1]} \varphi (y(x))= \min_{y \in D} \varphi(y).
\end{equation}

@@

\subsection{Solving the one-dimensional reduced optimization problem.} 
Using the dimensionality reduction results in one more additional advantage of the proposed approach: many well-known one-dimensional global search algorithms (possibly, after some additional generalisation) can be used to solve the initial multidimensional MOO problem from (\ref{eq:01}) \cite{c22,c23,c24,c25,c26,c27,c28,c29}. At the same time, it should be noted that most works where dimensionality reduction is a key feature for solving multiextremal problems like (\ref{eq:04}) rely on the information-statistical theory of global search \cite{c15}. This theory has provided the a basis for developing a large number of efficient methods for multi-extremal optimization \cite{c28,c29,c30,c31,c32,c33,c34,c35,c36,c37}.
